---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggmemo

<!-- badges: start -->
<!-- badges: end -->

Need to label a data point or show percent change on a ggplot2 chart?
ggmemo does it in one line — no manual coordinate math, no fiddly
arrow positioning.

| You want to... | Use |
|---|---|
| Point at a data point with an arrow and label | `annotate_callout(data, where = year == 2024, label = "Peak")` |
| Show percent change between two rows | `annotate_change(data, from = ..., to = ..., value = sales)` |
| Show absolute difference | `annotate_change(..., format = "absolute")` |
| Show change in percentage points | `annotate_change(..., format = "points")` |
| Use custom colors | `annotate_change(..., colors = c(up = "#1B9E77", down = "#D95F02", flat = "#999"))` |

## Installation

``` r
# install.packages("pak")
pak::pak("lindsay-lintelman/ggmemo")
```

## Examples

### Callout annotation

Point at a specific data row with an arrow and label — one line of ggmemo
code instead of manual arrow and label coordinates:

```{r callout, message = FALSE}
library(ggplot2)
library(ggmemo)

ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line() +
  annotate_callout(
    economics,
    where = date == as.Date("2009-10-01"),
    label = "Peak unemployment",
    position = "bottom-left"
  ) +
  labs(
    title = "U.S. Unemployment (thousands)",
    x = NULL, y = NULL
  )
```

### Change annotation

Show the delta between two data points with a color-coded arrow and
auto-formatted label:

```{r change, message = FALSE}
quarterly_revenue <- data.frame(
  quarter = factor(c("Q1", "Q2", "Q3", "Q4"),
                   levels = c("Q1", "Q2", "Q3", "Q4")),
  revenue = c(120, 145, 132, 158)
)

ggplot(quarterly_revenue, aes(x = quarter, y = revenue)) +
  geom_col(fill = "grey70", width = 0.6) +
  annotate_change(
    quarterly_revenue,
    from = quarter == "Q1",
    to = quarter == "Q4",
    value = revenue
  ) +
  labs(
    title = "Quarterly Revenue ($K)",
    x = NULL, y = NULL
  )
```
