---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

# ggmemo

<!-- badges: start -->
[![R-CMD-check](https://github.com/lindsay-lintelman/ggmemo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lindsay-lintelman/ggmemo/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Add arrows, labels, and change annotations to ggplot2 business charts
in one line of code.

```{r hero, echo = FALSE, message = FALSE}
library(ggplot2)
library(ggmemo)

econ <- economics
econ$unemp_rate <- econ$unemploy / econ$pop * 100

ggplot(econ, aes(x = date, y = unemp_rate)) +
  geom_line() +
  annotate_callout(
    econ,
    where = date == as.Date("1982-12-01"),
    label = "Volcker recession",
    position = "top-right",
    nudge = c(-1500, 0.5),
    arrow = NULL,
    point.padding = 2,
    alpha = .7
  ) +
  annotate_callout(
    econ,
    where = date == as.Date("2009-10-01"),
    label = "'08 Financial Crisis",
    position = "bottom-right",
    nudge = c(-2500, 0.5),
    arrow = NULL,
    point.padding = 2,
    alpha = .7
  ) +
  annotate_change(
    econ,
    from = date == as.Date("1982-12-01"),
    to = date == as.Date("2009-10-01"),
    value = unemp_rate,
    arrow_size = .08,
    arrow_type = "closed",
    curvature = -.3,
    format = "points"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(title = "U.S. Unemployment as Share of Population (%)",
       x = NULL, y = NULL) +
  theme(plot.margin = margin(10, 60, 10, 10))
```

## Overview

| You want to... | Use |
|---|---|
| Point at a data point with an arrow and label | `annotate_callout(data, where = year == 2024, label = "Peak")` |
| Show percent change between two rows | `annotate_change(data, from = ..., to = ..., value = sales)` |
| Show absolute difference | `annotate_change(..., format = "absolute")` |
| Show change in percentage points | `annotate_change(..., format = "points")` |
| Use custom colors | `annotate_change(..., colors = c(up = "#1B9E77", down = "#D95F02", flat = "#999"))` |

Without ggmemo, annotating a ggplot2 chart means hardcoding
coordinates, computing deltas, formatting labels, and choosing
colors — roughly 10 lines of manual work per annotation. ggmemo
replaces that with a single function call.

## Installation

```r
# install.packages("pak")
pak::pak("lindsay-lintelman/ggmemo")
```

## Examples

*Charts below are rendered at 7 × 4.5 in, 150 dpi.*

### Callout annotation

Point at a specific data row with an arrow and label:

```{r callout, message = FALSE}
library(ggplot2)
library(ggmemo)

# fig.width = 7, fig.height = 4.5, dpi = 150
econ <- economics
econ$unemp_rate <- econ$unemploy / econ$pop * 100

ggplot(econ, aes(x = date, y = unemp_rate)) +
  geom_line() +
  annotate_callout(
    econ,
    where = date == as.Date("1982-12-01"),
    label = "Volcker recession",
    position = "top-right",
    nudge = c(-1500, 0.5),
    arrow = NULL,
    point.padding = 2,
    alpha = .7
  ) +
  annotate_callout(
    econ,
    where = date == as.Date("2009-10-01"),
    label = "'08 Financial Crisis",
    position = "bottom-right",
    nudge = c(-2500, 0.5),
    arrow = NULL,
    point.padding = 2,
    alpha = .7
  ) +
  annotate_change(
    econ,
    from = date == as.Date("1982-12-01"),
    to = date == as.Date("2009-10-01"),
    value = unemp_rate,
    arrow_size = .08,
    arrow_type = "closed",
    curvature = -.3,
    format = "points"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(title = "U.S. Unemployment as Share of Population (%)",
       x = NULL, y = NULL) +
  theme(plot.margin = margin(10, 60, 10, 10))
```

### Time series change

Show the delta between two data points with a color-coded arrow:

```{r time-series-change, message = FALSE}
# fig.width = 7, fig.height = 4.5, dpi = 150
econ <- economics
econ$unemp_rate <- econ$unemploy / econ$pop * 100

ggplot(econ, aes(x = date, y = unemp_rate)) +
  geom_line(colour = "grey40") +
  annotate_change(
    econ,
    from = date == as.Date("2009-10-01"),
    to = date == as.Date("2015-03-01"),
    value = unemp_rate,
    arrow_pad = 0.05,
    curvature = -.35,
    arrow_size = 0.08,
    x = as.Date("2015-01-01"),
    y = 4.95,
    fill = NA
  ) +
  labs(title = "U.S. Unemployment as Share of Population (%)",
       x = NULL, y = NULL) +
  theme_minimal()
```

### Change annotation

```{r change, message = FALSE}
# fig.width = 7, fig.height = 4.5, dpi = 150
quarterly_revenue <- data.frame(
  quarter = factor(c("Q1", "Q2", "Q3", "Q4"),
                   levels = c("Q1", "Q2", "Q3", "Q4")),
  revenue = c(120, 145, 132, 158)
)

ggplot(quarterly_revenue, aes(x = quarter, y = revenue)) +
  geom_col(fill = "steelblue", width = 0.6) +
  annotate_change(
    quarterly_revenue,
    from = quarter == "Q3",
    to = quarter == "Q4",
    value = revenue,
    arrow_pad = .2,
    curvature = -.3,
    arrow_type = "closed",
    y = 180,
    fill = NA
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.4))) +
  labs(title = "Quarterly Revenue ($K)", x = NULL, y = NULL) +
  theme_minimal(base_size = 13)
```

## Learning more

* `vignette("narrating-business-charts")` — full walkthrough with
  customization, multiple annotations, and common mistakes.
* `?annotate_callout` and `?annotate_change` — function reference
  with all arguments and examples.

## Related packages

ggmemo is focused on callout and change annotations for business
charts. For other annotation needs:

* [ggpp](https://docs.r4photobiology.info/ggpp/) — precise
  annotation positioning with NPC coordinates. ggmemo is built on
  ggpp.
* [ggrepel](https://ggrepel.slowkow.com/) — automatically
  reposition text labels to avoid overlaps.
* [ggannotate](https://github.com/MattCowgill/ggannotate) —
  interactively annotate plots in RStudio.

## Code of Conduct

Please note that the ggmemo project is released with a [Contributor Code of
Conduct](https://lindsay-lintelman.github.io/ggmemo/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
