---
title: "Narrating Business Charts with ggmemo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Narrating Business Charts with ggmemo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

<!-- DRAFT — rewrite in your voice -->

A ggplot2 chart shows data. An *annotated* chart tells a story. ggmemo
adds the storytelling layer: arrows that point at key moments, labels
that name them, and change annotations that quantify what happened.

This vignette walks through a realistic example — quarterly revenue
data — from a bare chart to a narrated one, introducing each ggmemo
function along the way.

## The data

```{r setup}
library(ggplot2)
library(ggmemo)

revenue <- data.frame(
  quarter = factor(c("Q1", "Q2", "Q3", "Q4"),
                   levels = c("Q1", "Q2", "Q3", "Q4")),
  revenue = c(120, 145, 132, 158)
)
```

## A chart without annotations

<!-- DRAFT -->

Start with a plain bar chart. It's accurate, but it doesn't tell the
reader what to notice. Which quarter matters? Is the trend good or bad?
The viewer has to do the work.

```{r bare-chart}
p <- ggplot(revenue, aes(x = quarter, y = revenue)) +
  geom_col(fill = "steelblue", width = 0.6) +
  labs(title = "2024 Quarterly Revenue ($K)", x = NULL, y = NULL) +
  theme_minimal()
p
```

## Calling out a key data point

<!-- DRAFT -->

`annotate_callout()` points at a specific row in your data with an
arrow and label. You tell it *which* row with a filter expression
(the same syntax as `dplyr::filter()`), and it handles the arrow,
the box, and the positioning.

```{r callout}
p +
  annotate_callout(
    revenue,
    where = quarter == "Q4",
    label = "Record quarter",
    position = "top-left"
  )
```

The `position` argument controls which direction the label sits
relative to the data point: `"top-right"` (default), `"top-left"`,
`"bottom-right"`, or `"bottom-left"`.

## Showing the change between two points

<!-- DRAFT -->

`annotate_change()` draws a color-coded arrow between two rows and
labels the midpoint with the computed delta. Green for increases, red
for decreases.

```{r change-percent}
p +
  annotate_change(
    revenue,
    from = quarter == "Q1",
    to = quarter == "Q4",
    value = revenue
  )
```

The `format` argument controls how the delta is displayed:

```{r change-absolute}
# Absolute difference
p +
  annotate_change(
    revenue,
    from = quarter == "Q1",
    to = quarter == "Q4",
    value = revenue,
    format = "absolute"
  )
```

```{r change-points}
# Percentage points — useful when the data is already a rate
rates <- data.frame(
  year = 2020:2023,
  rate = c(3.5, 8.1, 5.4, 3.7)
)

ggplot(rates, aes(x = year, y = rate)) +
  geom_line() +
  geom_point() +
  annotate_change(
    rates,
    from = year == 2020,
    to = year == 2021,
    value = rate,
    format = "points"
  ) +
  labs(title = "Unemployment Rate", y = "Rate (%)") +
  theme_minimal()
```

## Putting it all together

<!-- DRAFT -->

The real power is combining both annotations on one chart. The callout
names the moment; the change quantifies it.

```{r combined}
ggplot(revenue, aes(x = quarter, y = revenue)) +
  geom_col(fill = "steelblue", width = 0.6) +
  annotate_callout(
    revenue,
    where = quarter == "Q4",
    label = "Record quarter",
    position = "top-left"
  ) +
  annotate_change(
    revenue,
    from = quarter == "Q1",
    to = quarter == "Q4",
    value = revenue
  ) +
  labs(title = "2024 Quarterly Revenue ($K)", x = NULL, y = NULL) +
  theme_minimal()
```

### Multiple change annotations

You can add more than one `annotate_change()` to show
quarter-over-quarter movement:

```{r multiple}
ggplot(revenue, aes(x = quarter, y = revenue)) +
  geom_col(fill = "grey70", width = 0.6) +
  annotate_change(revenue, from = quarter == "Q1",
                  to = quarter == "Q2", value = revenue) +
  annotate_change(revenue, from = quarter == "Q2",
                  to = quarter == "Q3", value = revenue) +
  annotate_change(revenue, from = quarter == "Q3",
                  to = quarter == "Q4", value = revenue) +
  labs(title = "Quarter-over-Quarter Changes", x = NULL, y = NULL) +
  theme_minimal()
```

### Time series example

<!-- DRAFT -->

ggmemo works with Date x-axes too. Here's a savings rate time series
with a callout at the all-time low and a change annotation showing
the recovery:

```{r time-series}
ggplot(economics, aes(x = date, y = psavert)) +
  geom_line(colour = "grey40") +
  annotate_callout(
    economics,
    where = date == as.Date("2005-07-01"),
    label = "All-time low",
    nudge = c(365, 1)
  ) +
  annotate_change(
    economics,
    from = date == as.Date("2005-07-01"),
    to = date == as.Date("2012-12-01"),
    value = psavert,
    format = "points"
  ) +
  labs(
    title = "U.S. Personal Savings Rate",
    subtitle = "Recovery after the 2005 low",
    x = NULL, y = "Savings rate (%)"
  ) +
  theme_minimal()
```

## Customization

### Colors

<!-- DRAFT -->

By default, `annotate_change()` uses dark green for increases, dark
red for decreases, and grey for no change. Override with the `colors`
argument:

```{r custom-colors}
p +
  annotate_change(
    revenue,
    from = quarter == "Q1",
    to = quarter == "Q4",
    value = revenue,
    colors = c(up = "#1B9E77", down = "#D95F02", flat = "#7570B3")
  )
```

### Label styling

Both functions accept `...` arguments that pass through to the
underlying ggplot2 layer. Common overrides:

```{r custom-style}
p +
  annotate_callout(
    revenue,
    where = quarter == "Q4",
    label = "Record quarter",
    position = "top-left",
    size = 5,
    fill = "lightyellow",
    colour = "grey30"
  )
```

### Nudge

<!-- DRAFT -->

`annotate_callout()` auto-computes how far to offset the label from
the data point. When the heuristic doesn't produce the right result
(common with wide data frames that have many numeric columns), set
`nudge` explicitly:

```{r nudge}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line() +
  annotate_callout(
    economics,
    where = date == as.Date("2009-10-01"),
    label = "Peak unemployment",
    nudge = c(800, 1000)
  ) +
  theme_minimal()
```

## Common mistakes

<!-- DRAFT — these came from dogfooding sessions -->

### Character columns need `factor()`

If your x-axis column is character (common after `read.csv()`),
`annotate_change()` will tell you to convert it. Always specify
`levels` to preserve data order — plain `factor()` sorts
alphabetically:

```r
# Wrong — months sort alphabetically (Apr, Feb, Jan, ...)
data$month <- factor(data$month)

# Right — preserves the order in your data
data$month <- factor(data$month, levels = unique(data$month))
```

### Date-like strings need `as.Date()`

If your CSV has dates as strings like `"2024-01-15"`, convert them
before plotting:

```r
data$date <- as.Date(data$date)
```

### Use `colour`, not `color`, in `...`

ggplot2 uses British spelling. American `color` works but may trigger
a "Duplicated aesthetics" warning when the function already sets a
default `colour`. Prefer `colour` to be safe:

```r
annotate_callout(..., colour = "red")
```

### `size` is text size, not label box size

To make annotation text larger, use `size` (in mm). To adjust the
box padding, use `label.padding`:

```r
annotate_callout(..., size = 5, label.padding = unit(0.4, "lines"))
```

## What ggmemo doesn't do

<!-- DRAFT -->

ggmemo is focused on two specific tasks: callout annotations and
change annotations. For other annotation needs:

- **Avoiding label overlaps**: use the
  [ggrepel](https://ggrepel.slowkow.com/) package.
- **Precise label positioning with NPC coordinates**: use the
  [ggpp](https://docs.r4photobiology.info/ggpp/) package (which
  ggmemo is built on).
- **Interactive annotations** (hover, click): use
  [plotly](https://plotly.com/r/) or
  [ggiraph](https://davidgohel.github.io/ggiraph/).
- **Themes and styling**: use ggthemes, hrbrthemes, or bbplot.
